---
title: "Drug Trial Analysis"
subtitle: ""
author: "Prachi Patel"
date: "2022-10-14"
excerpt: " This project analyzes data from a drug trial with 500 patients to test a new medication that aims to increase white blood cell count. We compare results between a treatment group and a control group to see if the drug works. The study also looks at how other factors like heart rate and blood pressure affect white blood cell counts."
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options: 
  markdown: 
    wrap: sentence

---

### Introduction

This project analyzes data from a drug trial with 500 patients to test a new medication that aims to increase white blood cell count. We compare results between a treatment group and a control group to see if the drug works. The study also looks at how other factors like heart rate and blood pressure affect white blood cell counts.

# Data Dictionary

```{r echo=FALSE, results='asis'}
options(knitr.kable.NA = '')
data.frame(Variable = c("ID","gender","HR","BP","Wgt1","Wgt2","Exercise","Hgt","Smoke","Drink","Group","WBC","Income"),
           Description = c("UserID (unique to each respondent)","Gender of patient","Heart Rate (Low, Normal, High)","Blood Pressure (Low, Normal, High)","Patient Weight 1 week before test start","Patient Weight at test start","Minutes per week patient exercises","Height of patient in inches","Does the patient smoke?","Does the patient drink alcohol?","Test/Control","White Blood Cell Count","Annual Income")) |>
  knitr::kable()
```

### Initial Setup

This section is for the basic set up.
It will clear all the plots, the console and the workspace.
It also sets the overall format for numbers.

```{r echo=T, results='hide'}
if(!is.null(dev.list())) dev.off()
cat("\014") 
rm(list=ls())
options(scipen=9)
```

### Load Packages

This section loads and attaches all the necessary packages.

```{r message=FALSE, warning=FALSE}
if(!require(readxl)){install.packages("readxl")}
library("readxl")

if(!require(pastecs)){install.packages("pastecs")}
library("pastecs")

if(!require(lattice)){install.packages("lattice")}
library("lattice")

if(!require(ggplot2)){install.packages("ggplot2")}
library("ggplot2")

if(!require(tinytex)){install.packages("tinytex")}
library("tinytex")

```

### Import Data

```{r results='hide'}
ExcelFile <- read_excel("Drug_Trial_Data.xlsx")
ExcelFile <- as.data.frame(ExcelFile)
head(ExcelFile)
```

### Data Transformation and Preparation

> Initial Transformation

Transforming character variables to factor variables.
  Using factor function we can convert character variable to factor variable.
        
```{r}
ExcelFile$gender <- as.factor(ExcelFile$gender)
ExcelFile$HR <- as.factor(ExcelFile$HR)
ExcelFile$BP <- as.factor(ExcelFile$BP)
ExcelFile$Smoke <- as.factor(ExcelFile$Smoke)
ExcelFile$Drink <- as.factor(ExcelFile$Drink)
ExcelFile$Group <- as.factor(ExcelFile$Group)
head(ExcelFile)
```

> Reduce Dimensionality
  
Applying the Missing Value Filter to remove appropriate columns of data using summary function, summarized all data in the data source. 

Based on the summary data 492 values out of 500 is blank or null for Income field.
    
```{r}
summary(ExcelFile)
ExcelFile <- ExcelFile[-c(13)]
summary(ExcelFile)
```
  
Applying the Low Variance Filter to remove appropriate columns of data.

Removed Hgt column because Height's coef.var value is 0.04.
      
```{r}
stat.desc(ExcelFile)
summary(ExcelFile)
table(ExcelFile$Hgt)
ExcelFile <- ExcelFile[-c(8)]
stat.desc(ExcelFile)
head(ExcelFile)
```

Applying the High Correlation Filter to remove appropriate columns of data.

Removing Wgt2 fields because Wgt2's correlation is 0.99; which is almost same as 1 for Wgt1.
    
```{r}
cor(ExcelFile[sapply(ExcelFile, function(x) !is.factor(x))])
ExcelFile <- ExcelFile[-c(6)]
head(ExcelFile)
```
    
Dropping the ID and Gender fields because both columns data is not useful for the analytics. 

As ID is the row number for each field and for gender column all data are marked as Female.

```{r}
ExcelFile <- ExcelFile[-c(1:2)]
head(ExcelFile)
```

> Outliers
  
Creating box plots of all relevant variables (i.e. numeric, non-binary) to determine outliers.

Created box plot for Weight, Exercise, and White Blood Cell fields as remaining fields are non-numeric.
      
```{r}
#Weight Data

boxplot(ExcelFile$Wgt1, horizontal=TRUE, col=c("#6bc9c2"), pch=20, main = "Weight - Box Plot")
densityplot( ~ ExcelFile$Wgt1, pch=6,col=c("#6bc9c2"), xlab = "Weight")

#Exercise Data
boxplot(ExcelFile$Exercise, horizontal=TRUE,col=c("#6bc9c2"), pch=20, main = "Exercise - Box Plot")
densityplot( ~ ExcelFile$Exercise, pch=6,col=c("#6bc9c2"), xlab = "Exercise")


boxplot(ExcelFile$WBC, horizontal=TRUE,col=c("#6bc9c2"), pch=20, main = "White Blood Cell - Box Plot")
densityplot( ~ ExcelFile$WBC, pch=6,col=c("#6bc9c2"), xlab = "White Blood Cell")

```
    
Commenting on any outliers you see and deal with them appropriately

Based on the previous box plots and density plots for Weight, Exercise and White Blood Cells fields; Weight and White Blood Cell fields has outlines.

For Weight, we have removed outliners data where value is less than 58.

For White blood cell, we have removed outliners data where value is less than 4100 and greater than 9700.
         
```{r}

#Code to delete outliner data where Weight value is less than 58.
nr <- which(ExcelFile$Wgt1 <58)
ExcelFile <- ExcelFile[-c(nr),]
boxplot(ExcelFile$Wgt1, horizontal=TRUE, col=c("#e8bbfa"), pch=20, main = "Weight - Box Plot")
densityplot( ~ ExcelFile$Wgt1, pch=6,col=c("#e8bbfa"), xlab = "Weight")

#Code to delete outliner data where WBC value is less than 4100 and greater than 9700.
nr <- which(ExcelFile$WBC < 4100)
ExcelFile <- ExcelFile[-c(nr),]
nr <- which(ExcelFile$WBC > 9700)
ExcelFile <- ExcelFile[-c(nr),]
boxplot(ExcelFile$WBC, horizontal=TRUE, col=c("#e8bbfa"), pch=20, main = "White Blood Cell - Box Plot")
densityplot( ~ ExcelFile$WBC, pch=6,col=c("#e8bbfa"), xlab = "White Blood Cell")
```
    
### Organizing Data

> Scatter Plots

Created a histogram for one of the Weight variables.
    
```{r}
histogram(~Wgt1, dat=ExcelFile, breaks=10,col = c("#6bc9c2"), type="density", xlab="Weight", main="Histogram of Weight")
```
    
Created a histogram for Exercise. 
    
```{r}
histogram(~Exercise, dat=ExcelFile, breaks=10,col = c("#6bc9c2"), type="density", xlab="Exercise", main="Histogram of Exercise")
```

Created a scatter plot showing the relationship between Exercise and Weight. (note: Exercise should be on the x-axis, Weight should be the y-axis)
    
```{r}
plot(Wgt1 ~ Exercise,
     data=ExcelFile,
     col = c("#6bc9c2"),
     pch=25,
     main="Scatterplot - Exercise & Weight",
     xlab="Exercise",
     ylab="Weight")
abline(0, 1, col='blue')
```

> Conclusion from the chart...

Based on the Scatter plot graph, we can say that exercise and weight data is not in a direct relationship as all data is show all over the graph.
      
Calculating a correlation coefficient between these two variables. 

Based on the correlation coefficient value, we can say that weight and exercise value are correlated to each other.
    
```{r}
ExcelFile$Wgt1 <- as.numeric(ExcelFile$Wgt1)
ExcelFile$Exercise <- as.numeric(ExcelFile$Exercise)
cor.test(ExcelFile$Wgt1,ExcelFile$Exercise, method="spearman")
```
    
### Inference

> Normality

Created a QQ Normal plot of White Blood Cell counts.
    
```{r}
qqnorm(ExcelFile$WBC, pch = 1,col = c("#6bc9c2"), frame = FALSE, main = "Normal Q-Q Plot for White Blood Cell")
qqline(ExcelFile$WBC, col = "black", lwd = 2)
```
    
Conducted a statistical test for normality on White Blood Cell counts.

For the statistical test, we can perform shapiro test to get the value for normality.

For this, p-value is 0.1515. We can say that data is normally distributed.
    
```{r}
shapiro.test(ExcelFile$WBC)
```

    
Now let's check if is White Blood Cell count normally distributed? What led you to this conclusion?

Based on the p-value get using shapiro test and the Q-Q plot graph, we can say that White Blood Cell data is normally distributed.
    
> Statistically Significant Differences

Compared White Blood Cell counts between the treatment and control group using a suitable hypothesis test.

Below is the process to choose comparison test.
    
```{r}
#As data is in the continuous manner, we will compare the means and groups if there are two or more than two and whether data meets the main three assumptions or not. 

#Data is independent.

#Data is normally distributed.
## Using Shapiro test p-value = 0.1515, we can that data is normally distributed.  
shapiro.test(ExcelFile$WBC)

#Variance is unknown.
##Based on f-test p-value = 0.6327, we can say that variances are same.
res.ftest <- var.test(WBC ~ Group, data = ExcelFile)
res.ftest

#Data met all the assumptions, we are performing t-test to mean.
res <- t.test(WBC ~ Group, data = ExcelFile, var.equal = TRUE)
res
```
    
Based on previous assumptions conditions, all 3 conditions are satisfied to perform the t-test.

Based on p-value for WBC counts, we can say that means are different. 

For T-test, if p > 0.05 - Means are same.
            if p < 0.05 - Means are different.
    
```{r}
t.test(WBC ~ Group, data = ExcelFile, var.equal = TRUE)

bwplot(WBC ~ Group, data=ExcelFile, 
       main="WBC by Group",
       xlab="Group", ylab = "WBC",  pch = '|')

```
    
> Multiple Statistical Differences

Determining if White Blood Cell count varies by Heart Rate Level using ANOVA (statistical) and a sequence of boxplots (graphical).
  
Using ANOVA test result, where pr = 0.768 we can say that means are same.
    
```{r}
summary(aov(WBC ~ HR, data=ExcelFile))

boxplot(WBC ~ HR, data=ExcelFile,
main="WBC by HR level",
ylab = "WBC",
xlab = "HR",
col=c("#6bc9c2"),
range=0)

```
    
Determining if White Blood Cell count varies by Blood Pressure Level using ANOVA and a sequence of boxplots.

Using ANOVA test result, where pr = 0.322 we can say that means are same.
    
```{r}
summary(aov(WBC ~ BP, data=ExcelFile))

boxplot(WBC ~ BP, data=ExcelFile,
main="WBC by BP Level",
xlab = "BP",
ylab = "WBC",
col=c("#6bc9c2"),
range=0)
```
