---
title: "Cancer Diagnosis Prediction Model"
author: "Prachi Patel"
date: "2022-12-13"
excerpt: "This project aims to predict cancer diagnosis using patient data. It compares different methods like Logistic Regression, Naive Bayes, and Linear Discriminant Analysis to find the most accurate model. The goal is to identify the best way to predict cancer while minimizing errors."
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

### Introduction

This project aims to predict cancer diagnosis using patient data. It compares different methods like Logistic Regression, Naive Bayes, and Linear Discriminant Analysis to find the most accurate model. The goal is to identify the best way to predict cancer while minimizing errors.

### Data Dictionary

```{r echo=FALSE, results='asis'}
options(knitr.kable.NA = '')
data.frame(Variable = c("Cancer","Age","HW","Hst","Exe","Smk","Drk","Hlth"),
           Description = c("The patient was diagnosed with cancer (0=no, 1=yes)","Age in years","A ratio of Height to Weight","History of Cancer in Family (0=No, 1=Yes)","Time spent exercising each week (in minutes)","Patient is smoker (0=No, 1=Yes)","Patient drinks alcohol (0=No, 1=Yes)","General health of patient. Five point scale: (VP) very poor health, (P) poor health, (A) average health, (G) good, (VG) very good health")) |>
  knitr::kable()
```

#### Initial Setup

This section is for the basic set up.
It will clear all the plots, the console and the workspace.
It also sets the overall format for numbers.

```{r echo=T, results='hide'}
if(!is.null(dev.list())) dev.off()
cat("\014") 
rm(list=ls())
options(scipen=9)
```

#### Load Packages

This section loads and attaches all the necessary packages.

```{r message=FALSE, warning=FALSE}
if(!require(readxl)){install.packages("readxl")}
library("readxl")

if(!require(pastecs)){install.packages("pastecs")}
library("pastecs")

if(!require(lattice)){install.packages("lattice")}
library("lattice")

if(!require(ggplot2)){install.packages("ggplot2")}
library("ggplot2")

if(!require(tinytex)){install.packages("tinytex")}
library("tinytex")

if(!require(corrgram)){install.packages("corrgram")}
library("corrgram")

if(!require(psych)){install.packages("psych")}
library("psych")

if(!require(partykit)){install.packages("partykit")}
library("partykit")

if(!require(klaR)){install.packages("klaR")}
  library("klaR")

if(!require(xlsx)){install.packages("xlsx")}
  library(xlsx)

```

#### Import Data

This code chunk is to read in the excel file and convert it into a data frame.

```{r message=FALSE, results='hide'}
ExcelFile <- read_excel("PROG8430_Final_22F_train.xlsx")
ExcelFile <- as.data.frame(ExcelFile)
head(ExcelFile)
```

#### Data Manupulation

Convert all the Character variable into Factor.

```{r results='hide'}
as.data.frame(unclass(ExcelFile),stringsAsFactors = TRUE) 
```

Converting all fields in numeric data type as we are checking correlation between all the fields.

```{r}
ExcelFile$Cancer <- as.numeric(ExcelFile$Cancer)
ExcelFile$Hst <- as.numeric(ExcelFile$Hst)
ExcelFile$Smk <- as.numeric(ExcelFile$Smk)
ExcelFile$Drk <- as.numeric(ExcelFile$Drk)
ExcelFile1 <- ExcelFile[c(1:7)]
cor(ExcelFile1[,unlist(lapply(ExcelFile1,is.numeric))], method="spearman")
```

Based on the correlation, we can say that there is strong correlation between Smoke and Drink in the model.

```{r}
ExcelFile1 <- ExcelFile1[-c(7)]
ExcelFile <- ExcelFile[-c(7)]
summary(ExcelFile)
```

> Checking Dimensionality reduction.

Using summary function, I have summarized all data in the data source.

```{r}
summary(ExcelFile)
```

Based on summary, we can say that no null data is available as no data is available in `nbr.null` row.

```{r}
stat.desc(ExcelFile)
```

> Checking Outliners

Create `barplot` for cancer field that shows if the patient was diagnosed with cancer or not.

This plot will show if any outliners are available or not.

```{r}
Cancer <- table(ExcelFile$Cancer)
Cancer <- Cancer[order(Cancer,decreasing=TRUE)]
barplot(Cancer,density = 30, angle = 45, main="The patient was diagnosed with cancer (0=no, 1=yes)", xlab="The patient was diagnosed with cancer", ylab = "Frequency", col=c("#6bc9c2"))
```

As per the graph we can say that there are no outliners are available for the cancer field.

Created `boxplot` for Age (i.e. numeric, non-binary) to determine outliers.

```{r}
boxplot(ExcelFile$Age, horizontal=TRUE, col=c("#6bc9c2"), pch=20, main = "Age in years")
densityplot( ~ ExcelFile$Age, pch=6,col=c("#6bc9c2"), xlab = "Age")
```

As per the graph we can say that there are no outliners are available for the Age field.

Created `boxplot` for Height to Weight (i.e. numeric, non-binary) to determine outliers.

```{r}
boxplot(ExcelFile$HW, horizontal=TRUE, col=c("#6bc9c2"), pch=20, main = "A ratio of Height to Weight")
densityplot( ~ ExcelFile$HW, pch=6,col=c("#6bc9c2"), xlab = "Height to Weight")
```

As per the graph we can say that there are outliners are available for the height to weight ratio field.

This code to delete outliner data where Height to Weight value is more than 2.

```{r}
nr <- which(ExcelFile$HW > 2)
ExcelFile <- ExcelFile[-c(nr),]
boxplot(ExcelFile$HW, horizontal=TRUE, col=c("#e8bbfa"), pch=20, main = "A ratio of Height to Weight")
densityplot( ~ ExcelFile$HW, pch=6,col=c("#e8bbfa"), xlab = "Height to Weight")
```

Now we can see that no outliners are available for the height to weight field.

Created `boxplot` for History of Cancer in Family field to determine outliers.

```{r}
Hst <- table(ExcelFile$Hst)
Hst <- Hst[order(Hst,decreasing=TRUE)]
barplot(Hst,density = 30, angle = 45, main="History of Cancer in Family (0=No, 1=Yes)", xlab="History of Cancer", ylab = "Frequency", col=c("#6bc9c2"))
```

And we can see that no outliners are available in the History of Cancer in Family field.

Created `boxplot` for Exercise (i.e. numeric, non-binary) to determine outliers.

```{r}
boxplot(ExcelFile$Exe, horizontal=TRUE, col=c("#6bc9c2"), pch=20, main = "Time spent exercising each week (in minutes) - Box Plot")
densityplot( ~ ExcelFile$Exe, pch=6,col=c("#6bc9c2"), xlab = "Exercise")
```

No outliners are available in the exercise field as well.

Now, let's check the outliners for the smoker field.

```{r}
Smk <- table(ExcelFile$Smk)
Smk <- Smk[order(Smk,decreasing=TRUE)]
barplot(Smk,density = 30, angle = 45, main="Patient is smoker (0=No, 1=Yes) - Bar Plot", xlab="Smoker", ylab = "Frequency", col=c("#6bc9c2"))
```

And again we can see that no data available in smoker field as well.

Now, let's check the outliners for the health of the patient field as well.

```{r}
Hlth <- table(ExcelFile$Hlth)
Hlth <- Hlth[order(Hlth,decreasing=TRUE)]
barplot(Hlth,density = 30, angle = 45, main="General health of patient. Five point scale: (VP) very poor health, (P) poor 
health, (A) average health, (G) good, (VG) very good health - Bar Plot", xlab="Health", ylab = "Frequency", col=c("#6bc9c2"))

```

And for this field as well we can see that no outliners are available for this field as well.

> Full Model

Now, let's create a full model for all the variable with the Cancer variable as a dependent variable.

```{r}
full.model = glm(Cancer ~ . ,data=ExcelFile, na.action=na.omit, family="binomial")
summary(full.model)
pred <- predict(full.model, newdata=ExcelFile)
```

> Backward Model

Creating another model using backward selection.

```{r}
back.model = step(full.model, direction="backward", details=TRUE)
summary(back.model)
pred <- predict(back.model, newdata=ExcelFile)
```

-------------------------------------------------------------------------------

Let's analyse the backward model.

**AIC** - Backward model has lowest AIC value.

**Deviance** - Backward model has higher deviance difference.

**Residual symmetry** - Residual symmetry value is near by zero for backward model.

**z-values** - all Pr(>z) values less than 0.05 is consider as the better model.

For Backward model, Pr(>z) value is less than 0.05 for one and another variable Pr(>z) value is near by 0.05.

> Based on all these observations, we can say that *backward model* is the better model than other model. 

-------------------------------------------------------------------------

As backward model is better, let's create a Logistic Regression using Backward.

```{r}
# Converting these field into factor.
ExcelFile$Cancer <- as.factor(ExcelFile$Cancer)
ExcelFile$Hst <- as.factor(ExcelFile$Hst)
ExcelFile$Smk <- as.factor(ExcelFile$Smk)
```

In regression, the dependent variable is marked as Y and the independent variable is marked as X. 

So, here adding the Cancer as dependent variable and other variables as the independent variables.

```{r}
# glm = 'Generalized Linear Models'
glm.mod <- glm(Cancer ~ .,family="binomial", data=ExcelFile, na.action=na.omit)
stp.glm <- step(glm.mod)
```

Classifies

```{r}
resp <- predict(back.model, type="response")
head(resp,10)
```

adding a condition of more than 50% chance then 1 else 0.

```{r}
class <- ifelse(resp > 0.5,1,0)
head(class)
```

> Confusion Matrix for Logistic Regression – Backward

creating the confusion matrix.

```{r}
LR_CF <-table(ExcelFile$Cancer, class, dnn = list("Actual","Predicted"))
LR_CF

LR_TP <-LR_CF[2,2]
LR_TN <-LR_CF[1,1]
LR_FP <-LR_CF[1,2]
LR_FN <-LR_CF[2,1]
```

> Naive-Bayes Classification

In regression, the dependent variable is marked as Y and the independent variable is marked as X. 

So, here adding the OT as dependent variable and other variables as the independent variables.

```{r}
Naive <- NaiveBayes(Cancer ~ .,family="binomial", data=ExcelFile, na.action=na.omit)
```

> Confusion Matrix for Naive-Bayes Classification

Classifies

```{r warning=FALSE}
pred_bay <- predict(Naive,ExcelFile)
```

Creates Confusion Matrix

```{r}
Naive_CF <- table(Actual=ExcelFile$Cancer, Predicted=pred_bay$class)
Naive_CF

Naive_TP <-Naive_CF[2,2]
Naive_TN <-Naive_CF[1,1]
Naive_FP <-Naive_CF[1,2]
Naive_FN <-Naive_CF[2,1]
```

> Linear Discriminant Analysis

```{r}
LDA <- lda(Cancer ~ .,data = ExcelFile, na.action=na.omit)
```

> Confusion Matrix for Linear Discriminant Analysis

Classifies
```{r}
pred_dis <- predict(LDA, data=ExcelFile)
```

Confusion Matrix

```{r}
LDA_CF <- table(Actual=ExcelFile$Cancer, Predicted=pred_dis$class)
LDA_CF
```

Defined

```{r}
LDA_TP <-LDA_CF[2,2]
LDA_TN <-LDA_CF[1,1]
LDA_FP <-LDA_CF[1,2]
LDA_FN <-LDA_CF[2,1]
```

> Checking which classifier is most accurate.

Accuracy: The proportion of cases correctly classified: (TP+TN)/Total

Accuracy for LR classifier.

```{r}
LR_ACC <-(LR_TP+LR_TN)/sum(LR_CF)
cat("Logistic Regression: ")
LR_ACC
```

Accuracy for Naive classifier.

```{r}
Naive_ACC <-(Naive_TP+Naive_TN)/sum(Naive_CF)
cat("Naive Bayes Classification: ")
Naive_ACC
```

Accuracy for LDA Classifier.

```{r}
LDA_ACC <-(LDA_TP+LDA_TN)/sum(LDA_CF)
cat("Linear Discriminant Analysis: ")
LDA_ACC
```

Based on the accuracy values for all three defined classifiers, Linear Discriminant Analysis classifier is the most accurate 
because it's value is slightly higher than other two classifier's accuracy value.

> Checking which classifier minimizes false positives.

False Positive value for Logistic Regression Classifier.

```{r}
cat("Logistic Regression: ")
LR_FP
```

False Positive value for Naive Bayes Classification.

```{r}
cat("Naive Bayes Classification: ")
Naive_FP
```

False Positive value for Linear Discriminant Analysis Classifier

```{r}
cat("Linear Discriminant Analysis: ")
LDA_FP
```

Based on the FP(False Positive) values for all three classifiers, 
Linear Discriminant Analysis Classifier has the lowest value for false positives.

> Add Prediction

```{r}

Test <- read_excel("PROG8430_Final_22F_test.xlsx")
head(Test,10)
```



```{r}
# convert all the char variables to factor.
as.data.frame(unclass(Test),stringsAsFactors = TRUE) 

Test$Hst <- as.numeric(Test$Hst)
Test$Smk <- as.numeric(Test$Smk)

Test1 <-as.data.frame(unclass(Test),stringsAsFactors=TRUE)

# Predicting a cancer values for new data using backward model that I have created previously. 
Cancer <- predict(back.model, newdata=Test1)
head(Cancer)

Test1 <- cbind(Test,Cancer)

write.xlsx(Test1,"PROG8430_Final_22F_Pred.xlsx")

head(Test1)

```

